---
title: "Classification Part (vowel)"
author: "Andrei Kostin"
output: 
  pdf_document:
    #toc: true # Add table of contents
    #number_sections: true # Number sections
    latex_engine: xelatex
---

  Load the required libraries:

```{r}
library(readr)
library(tidyverse)
library(dplyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(caret)
library(class)
library(ggthemes)
library(corrplot)
library(MASS)
```

  Read the target dataset and name the columns. Variables F0–F9 represent acoustic formants and "Class" is the vowel to predict.

```{r}
df <- read.csv("../data/vowel.dat", 
               comment.char="@", header = FALSE)

names(df) <- c("TT", "SpeakerNumber", "Sex",
               "F0","F1", "F2", "F3", "F4", "F5", "F6", "F7", "F8", "F9", 
               "Class")
head(df)
```

  Convert numeric variables to categorical type:

```{r}
df$TT <- factor(df$TT)
df$SpeakerNumber <- factor(df$SpeakerNumber)
df$Sex <- factor(df$Sex)
# Recollapse Sex levels
df$Sex <- fct_collapse(df$Sex, 
             male = "0",
             female = "1")
df$Class <- factor(df$Class)

glimpse(df) # data structure
```
  ## 1. Classification with k-NN
  
  Normalize data for distance calculation:

```{r}
prepr <- preProcess(df[,4:13], method = "range")
df_scaled <- predict(prepr, df)
head(df_scaled)

```

  Define a function to train and evaluate k-NN on train/test files:

```{r}
# about return_data: if TRUE, also returns scaled data and predictions
# for plotting
run_knn_fold <- function(i, k = 5, tt = "test", return_data = FALSE) {
file <- paste0("../data/vowel-10-", i, "tra.dat") 
x_tra <- read.csv(file, comment.char="@", header=FALSE)
file <- paste0("../data/vowel-10-", i, "tst.dat") 
x_tst <- read.csv(file, comment.char="@", header=FALSE)
names(x_tra) <- c("TT", "SpeakerNumber", "Sex",
                  "F0","F1", "F2", "F3", "F4", "F5", "F6", "F7", "F8", "F9", 
                  "Class")
names(x_tst) <- names(x_tra)

x_tra$Class <- factor(x_tra$Class)
x_tst$Class <- factor(x_tst$Class)

# Normalize F0–F9
prepr <- preProcess(x_tra[,4:13], method = "range")
x_tra_scaled <- predict(prepr, x_tra[,4:13])
x_tst_scaled <- predict(prepr, x_tst[,4:13])

# Choose whether to evaluate on train or test
if (tt == "train") { 
  x_test <- x_tra_scaled
  y_test <- x_tra$Class
  }
else { 
  x_test <- x_tst_scaled 
  y_test <- x_tst$Class
  }

# Apply k-NN with scaled training data
knn.pred <- knn(x_tra_scaled, x_test, cl = x_tra$Class, k = k)


if (return_data){
  # Return both data and predictions for visualization
  val <- cbind(x_test, Class = y_test)
  list(data = val, pred = knn.pred)
  } else {
    # Calculate and return accuracy on the chosen set
    t <- table(knn.pred, y_test)
    #print(t)
    sum(diag(t)) / length(y_test)
  }


}

```

For example, use this function to calculate k-NN accuracy on test with k = 5 in the first fold:

```{r}
run_knn_fold(1, k = 5, tt = "test")
```
  Calculate the average k-NN accuracy (k = 5) across all folds:

```{r}
acc_knn_test <- mean(sapply(1:10, run_knn_fold, tt = "test"))
acc_knn_train <- mean(sapply(1:10, run_knn_fold, tt = "train"))
res_knn <- data.frame(model = "k-NN",
                      acc_train = acc_knn_train,
                      acc_test  = acc_knn_test)

res_knn
```

  Training accuracy is very high (nearly 1), test accuracy is slightly lower but still high, indicating good performance with slight overfitting.
  
  Visualization of k-NN predictions in fold 1: 

```{r}
res_plot <- run_knn_fold(1, k = 5, tt = "test", return_data = TRUE)
val <- as.data.frame(res_plot$data)
knn.pred <- res_plot$pred

# Unify true and predicted class levels to avoid factor issues
lev <- union(levels(factor(val$Class)), levels(factor(knn.pred)))
val$Class  <- factor(val$Class,  levels = lev)
knn.pred <- factor(knn.pred,  levels = lev)

# Prepare a long dataframe for F1 vs F2 plotting
plot_data <- val %>%
  mutate(prediction = knn.pred) %>%
  rename(known = Class) %>%
  gather(type, value, prediction, known)

ggplot(plot_data) +
  geom_point(aes(x = F1, y = F2, shape = type, color = value)) +
  labs(x = "Formant F1", y = "Formant F2" , 
       title = "F1 vs F2 (true values and predictions)") +
  scale_shape_manual(values = c(5, 3)) + 
  theme_minimal()
```
  This plot shows test points in the F1-F2 space, where diamonds correspond to true classes and crosses are k-NN predictions.
  Most crosses fall on diamonds of the same color, confirming high classification accuracy.

  Confusion matrix:

```{r}
cm <- confusionMatrix(knn.pred, val$Class)
cm
```
  Overall, k-NN achieves high accuracy with sensitivity and specificity close to 1 for most classes. Most errors are concentrated in class 6, which has smaller values compared to other classes.

  Combine all training files into one for robust evaluation and apply k-NN with caret:

```{r}
all_train <- lapply(1:10, function(i) {
  f <- paste0("../data/vowel-10-", i, "tra.dat")
  x <- read.csv(f, comment.char = "@", header = FALSE)
  names(x) <- c("TT","SpeakerNumber","Sex",
                "F0","F1","F2","F3","F4","F5","F6","F7","F8","F9",
                "Class")
  x
}) %>% bind_rows()

all_train$Class <- factor(all_train$Class)

suppressWarnings({
  knn.model <- train(Class ~ .-Class-TT-SpeakerNumber-Sex,
                     data = all_train,
                     method="knn", 
                     metric="Accuracy", 
                     tuneGrid = expand.grid(.k = 1:15))
})

knn.model
```
  With bootstrap on the combined training set, all k values between 1 and 15 show very high accuracy. This indicates that k-NN fits the available data very well, although this validation is based on resampling of the training set itself and therefore slightly overestimates real performance. The most suitable k values are from k = 2 to k = 5, offering a good compromise between high accuracy and moderate overfitting.

Visualization of the k-NN method:

```{r}
plot(knn.model)
```

  Calculate accuracy for different k values:

```{r}
k <- c(1, 3, 5, 7, 9, 11)
# For each k, calculate average accuracy on train and test  
get_acc_for_k <- function(k) {
  acc_test  <- mean(sapply(1:10, run_knn_fold, k = k, tt = "test"))
  acc_train <- mean(sapply(1:10, run_knn_fold, k = k, tt = "train"))
  c(acc_train = acc_train, acc_test = acc_test)
}

res_mat <- sapply(k, get_acc_for_k)

res_knn <- data.frame(k = k,
                      acc_train = res_mat["acc_train", ],
                      acc_test  = res_mat["acc_test", ])

res_knn
```

  Visualization of the results:
  
```{r}

plot_knn <- res_knn %>%
  gather(test_train, accuracy, acc_train, acc_test)

ggplot(plot_knn, aes(x = k, y = accuracy, color = test_train)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = k) + 
  labs(x = "k", y = "Mean Accuracy",
       title = "k-NN accuracy on training and test by k") +
  theme_minimal()
```
  As k increases, both accuracies decrease and the gap between train and test widens. This indicates that small k values give better performance.
  
  ## 2. Classification with LDA
  
  Define a function analogous to run_knn_fold but for LDA:
  
```{r}
run_lda_fold <- function(i, tt = "test", return_data = FALSE) {
file <- paste0("../data/vowel-10-", i, "tra.dat") 
x_tra <- read.csv(file, comment.char="@", header=FALSE)
file <- paste0("../data/vowel-10-", i, "tst.dat") 
x_tst <- read.csv(file, comment.char="@", header=FALSE)
names(x_tra) <- c("TT", "SpeakerNumber", "Sex",
                  "F0","F1", "F2", "F3", "F4", "F5", "F6", "F7", "F8", "F9", 
                  "Class")
names(x_tst) <- names(x_tra)

x_tra$Class <- factor(x_tra$Class)
x_tst$Class <- factor(x_tst$Class)

# Fit LDA model using only the training set
model <- lda(Class ~ .-TT-SpeakerNumber-Sex, data = x_tra)


if (tt == "train") { 
  test <- x_tra
  }
else { 
  test <- x_tst
  }

  # Get predicted classes and calculate accuracy
  pred <- predict(model, test)
  mean(pred$class == test$Class)

}

```

 Calculate average LDA accuracy across all folds:

```{r}
acc_lda_test <- mean(sapply(1:10, run_lda_fold, tt = "test"))
acc_lda_train <- mean(sapply(1:10, run_lda_fold, tt = "train"))
res_lda <- data.frame(model = "LDA",
                      acc_train = acc_lda_train,
                      acc_test  = acc_lda_test)

res_lda
```

  This analysis shows that LDA is a significantly less efficient classifier for the vowel dataset compared to k-NN, which achieves much higher results. There is not much difference between train and test values, so this model is also fairly stable, albeit with lower performance.
  
  LD1 and LD2 projection:

```{r}
lda_model <- lda(Class ~ .-TT-SpeakerNumber-Sex, data = all_train)

lda.data <- cbind(all_train, predict(lda_model)$x)

ggplot(lda.data, aes(x = LD1, y = LD2, color = Class)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "LDA projection: LD1 vs LD2") +
  theme_minimal()
```
  The reason for such low accuracy values is the poor data separability that can be observed graphically. The classes overlap heavily in space, resulting in the low LDA precision observed.

  Normality assessment for formants F1 and F2:
  
```{r}
ggplot(all_train, aes(x = F1)) +
  geom_histogram(bins = 20, color = "black", fill = "steelblue") +
  facet_wrap(~ Class, ncol = 4) +
  labs(title = "Histograms of F1 by class")
```
```{r}
ggplot(all_train, aes(x = F2)) +
  geom_histogram(bins = 20, color = "black", fill = "steelblue") +
  facet_wrap(~ Class, ncol = 4) +
  labs(title = "Histograms of F2 by class")
```
The histograms are mostly unimodal, though skewed. Multivariate normality holds only approximately. Since the dispersion clearly differs between classes, this violates the classical LDA assumption. 

  ## 3. Classification with QDA

  Define a function analogous to run_knn_fold and run_lda_fold but for QDA:

```{r}
run_qda_fold <- function(i, tt = "test", return_data = FALSE) {
file <- paste0("../data/vowel-10-", i, "tra.dat") 
x_tra <- read.csv(file, comment.char="@", header=FALSE)
file <- paste0("../data/vowel-10-", i, "tst.dat") 
x_tst <- read.csv(file, comment.char="@", header=FALSE)
names(x_tra) <- c("TT", "SpeakerNumber", "Sex",
                  "F0","F1", "F2", "F3", "F4", "F5", "F6", "F7", "F8", "F9", 
                  "Class")
names(x_tst) <- names(x_tra)

x_tra$Class <- factor(x_tra$Class)
x_tst$Class <- factor(x_tst$Class)

model <- qda(Class ~ .-TT-SpeakerNumber-Sex, data = x_tra)


if (tt == "train") { 
  test <- x_tra
  }
else { 
  test <- x_tst
  }
  
  pred <- predict(model, test)
  mean(pred$class == test$Class)
}
```

  Calculate average QDA accuracy across all folds:

```{r}
acc_qda_test <- mean(sapply(1:10, run_qda_fold, tt = "test"))
acc_qda_train <- mean(sapply(1:10, run_qda_fold, tt = "train"))
res_qda <- data.frame(model = "QDA",
                      acc_train = acc_qda_train,
                      acc_test  = acc_qda_test)

res_qda

```
  According to the results, QDA shows much better efficiency compared to LDA accuracy, although the method is still more efficient.

  Comparison of formant variances for three representative classes:
  
```{r}
features <- c("F0","F1","F2","F3","F4","F5","F6","F7","F8","F9")
cov0 <- cov(subset(all_train, Class == 0)[, features])
cov1 <- cov(subset(all_train, Class == 1)[, features])
cov2 <- cov(subset(all_train, Class == 2)[, features])

# calculate diagonals of covariance matrices
var_df <- data.frame(feature = features,
                     class0  = diag(cov0),
                     class1  = diag(cov1),
                     class2  = diag(cov2))

var_long <- var_df %>%
  pivot_longer(cols = c(class0, class1, class2),
               names_to = "class",
               values_to = "variance")

ggplot(var_long, aes(x = feature, y = variance, color = class, group = class)) +
  geom_line() +
  geom_point() +
  labs(title = "Variance comparison by class",
       x = "Formant", y = "Variance") +
  theme_minimal()

```
  The plot provides clear evidence that class dispersions are not equal. This explains why LDA showed such poor results.
  
  ## 4. Comparison of three algorithms
  
  Compare k-NN, LDA, and QDA results in a single dataframe:
  
```{r}
best_k <- 3   # best k for k-NN

# Calculate average k-NN accuracies
acc_knn_train <- mean(sapply(1:10, run_knn_fold, k = best_k, tt = "train"))
acc_knn_test  <- mean(sapply(1:10, run_knn_fold, k = best_k, tt = "test"))

res_all <- rbind(
  data.frame(model = "k-NN", acc_train = acc_knn_train, acc_test = acc_knn_test),
  data.frame(model = "LDA", acc_train = acc_lda_train, acc_test = acc_lda_test),
  data.frame(model = "QDA", acc_train = acc_qda_train, acc_test = acc_qda_test)
)

res_all
```

  Visualization of this comparison:

```{r}
ggplot(res_all, aes(x = model, y = acc_test, fill = model)) +
  geom_col() +
  ylim(0, 1) +
  labs(title = "Mean test accuracy by model",
       x = "Model", y = "Accuracy (test)") +
  theme_minimal()
```
  The analysis clearly shows that for the vowel dataset classification task, **k-NN** is the most efficient model, achieving the highest accuracy. This success is because vowel class boundaries are complex and non-linear. On the other hand, *LDA* performed poorly (60.1% accuracy) because the boundaries are not linear and the variances differ greatly. *QDA* improved accuracy but did not reach k-NN's performance.
  Consequently, the recommended final model for this task is **k-NN** with **k = 3**.