---
title: "Regression part (autoMPG6)"
author: "Andrei Kostin"
output: 
  pdf_document:
    #toc: true # Add table of contents
    #number_sections: true # Number sections
    latex_engine: xelatex
---

  Load the required libraries:

```{r}
library(readr)
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(MASS)
library(kknn)
```

  Read the target dataset and name the columns:

```{r}
df <- read.csv("../data/autoMPG6.dat", 
               comment.char="@", header = FALSE)

names(df) <- c("Displacement", "Horse_power", "Weight",
                 "Acceleration", "Model_year", "Mpg")
head(df) 
```

## 1. Linear Regression

  Analysis of all variables using quality metrics:

```{r}
fit_weight  <- lm(Mpg ~ Weight, data = df)
fit_hp      <- lm(Mpg ~ Horse_power, data = df)
fit_disp    <- lm(Mpg ~ Displacement, data = df)
fit_acc     <- lm(Mpg ~ Acceleration, data = df)
fit_year    <- lm(Mpg ~ Model_year, data = df)


simple_stats <- data.frame(var = c("Weight", "Horse_power", "Displacement", 
                                   "Acceleration", "Model_year"),
                           r2 = c(summary(fit_weight)$r.squared,
                                  summary(fit_hp)$r.squared,
                                  summary(fit_disp)$r.squared,
                                  summary(fit_acc)$r.squared,
                                  summary(fit_year)$r.squared),
                           r2adj = c(summary(fit_weight)$adj.r.squared,
                                     summary(fit_hp)$adj.r.squared,
                                     summary(fit_disp)$adj.r.squared,
                                     summary(fit_acc)$adj.r.squared,
                                     summary(fit_year)$adj.r.squared),
                           rmse  = c(sqrt(sum(fit_weight$residuals^2)/length(fit_weight$residuals)),
                                     sqrt(sum(fit_hp$residuals^2)/length(fit_hp$residuals)),
                                     sqrt(sum(fit_disp$residuals^2)/length(fit_disp$residuals)),
                                     sqrt(sum(fit_acc$residuals^2)/length(fit_acc$residuals)),
                                     sqrt(sum(fit_year$residuals^2)/length(fit_year$residuals)))
)

simple_stats
```
  The results of the five simple linear regressions show that *Weight* is the most informative predictor for *Mpg*, with an adjusted R2 of ~0.69 and the lowest RMSE of ~4.32.
*Displacement* and *Horse_power* also show strong relationships with *Mpg* (adjusted R2 of 0.65 and 0.60, respectively).
Therefore, the model *Mpg ~ Weight* is selected as the best simple linear regression model for this dataset.

  Scatterplot visualization of all predictors against Mpg:

```{r}
df_long <- df %>% pivot_longer(cols = -Mpg, names_to  = "regr", values_to = "values")

ggplot(df_long, aes(x = values, y = Mpg)) +
  geom_point(alpha = .4, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ regr, scales = "free_x") + 
  labs(title = "Scatterplots Mpg vs predictors", x = "predictors") + 
  theme_minimal()
```

  *Weight*, *Displacement*, and *Horse_power* show a strong negative linear relationship with *Mpg*: as these variables increase, *Mpg* decreases almost linearly.
  In contrast, the relationships of *Mpg* with *Acceleration* and especially *Model_year* are more scattered, making them less explanatory for regression.

## 2. Multiple Linear Regression

  Fit the full regression model with all variables:

```{r}
fit_all = lm(Mpg~., data = df)
summary(fit_all)
```
  According to the quality metrics, *Weight* and *Model_year* have highly significant coefficients, while *Displacement*, *Horse_power*, and *Acceleration* have high p-values, suggesting they add little additional information.
  Therefore, we remove variables one by one in order of relevance by p-value:

```{r}
fit1 = lm(Mpg~.-Horse_power, data = df)
summary(fit1)
```


```{r}
fit2 = lm(Mpg~.-Horse_power-Displacement, data = df)
summary(fit2)
```


```{r}
fit3 = lm(Mpg~.-Horse_power-Displacement-Acceleration, data = df)
summary(fit3)
```
  The results show a consistent decrease in residual error. The reduced model using only *Weight* and *Model_year* achieves an adjusted R2 of 0.807 and the lowest residual standard error. These variables align with the EDA hypotheses: greater weight decreases *Mpg*, while newer models have more efficient fuel consumption.
  
  Check interaction of the most relevant variables:
  
```{r}
fit4 = lm(Mpg~Weight*Model_year, data = df)
summary(fit4)
```
  Quality metrics improve substantially. This indicates that the effect of weight on *Mpg* depends on the model year.
  
  Adding non-linearity:

```{r}
fit5 = lm(Mpg~Weight + I(Weight^2) + Model_year, data = df)
summary(fit5)
```
  After including a quadratic term for *Weight*, the result improves considerably compared to all previous analyses.
  
  Combination of the last two steps:

```{r}
fit6 = lm(Mpg~Weight*Model_year + 
            Weight + I(Weight^2) + 
            Model_year + I(Model_year^2), data = df)
summary(fit6)
```
  The best results are obtained by combining interaction and quadratic terms for both *Weight* and *Model_year*. This confirms non-linear relationships and a weight effect that depends on model year.
  
  Plot Mpg vs Weight for each model year:

```{r}
ggplot(df, aes(x = Weight, y = Mpg)) +
  geom_point(alpha = .5, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ Model_year) +
  labs(title = "Mpg–Weight relationship by model year") +
  theme_minimal()
```

## 3.	k-NN for Non-parametric Regression

  The comparative plots show that **k-NN** predictions follow the global trends of real data reasonably well but underestimate extreme values:
 
```{r}
fitknn_all <- kknn(Mpg ~ ., df, df, k=7, distance=2, kernel="optimal", scale=TRUE)

df_long <- df %>% mutate(fitted = fitknn_all$fitted.values) %>% 
  pivot_longer(cols = -c(Mpg,fitted), names_to = "variable", values_to = "value")
df_long %>% ggplot(aes(x = value)) + 
  geom_point(aes(y = Mpg), alpha = .5) +
  geom_point(aes(y = fitted), color = "blue", shape = 20, size = 1.7) +
  facet_wrap(~ variable, scale="free") + theme_minimal()
```
  First, we read the train and test files and apply the linear model to obtain scores:

```{r}
# K-NN

name <- "autoMPG6"
run_lm_fold <- function(i, x, tt = "test") {
file <- paste("../data/", x, "-5-", i, "tra.dat", sep=""); x_tra <- read.csv(file, comment.char="@", header=FALSE)
file <- paste("../data/", x, "-5-", i, "tst.dat", sep=""); x_tst <- read.csv(file, comment.char="@", header=FALSE)
names(x_tra) <- c("Displacement", "Horse_power", "Weight",
                 "Acceleration", "Model_year", "Mpg")
names(x_tst) <- c("Displacement", "Horse_power", "Weight",
                 "Acceleration", "Model_year", "Mpg")
if (tt == "train") { 
  test <- x_tra 
  }
else { 
  test <- x_tst 
  }
fitMulti=lm(Mpg~Weight*Model_year + 
            Weight + I(Weight^2) + 
            Model_year + I(Model_year^2), x_tra)
yprime=predict(fitMulti,test)
sum(abs(test$Mpg-yprime)^2)/length(yprime) ##MSE
}
lmMSEtrain<-mean(sapply(1:5,run_lm_fold,name,"train"))
lmMSEtest<-mean(sapply(1:5,run_lm_fold,name,"test"))
cat("LM train score: ", lmMSEtrain, "\n")
cat("LM test score: ", lmMSEtest, "\n")
```
  Next, we run the same procedure with the k-NN model:

```{r}
library(kknn)
name <- "autoMPG6"
run_knn_fold <- function(i, x, tt = "test") {
file <- paste("../data/", x, "-5-", i, "tra.dat", sep=""); x_tra <- read.csv(file, comment.char="@", header=FALSE)
file <- paste("../data/", x, "-5-", i, "tst.dat", sep=""); x_tst <- read.csv(file, comment.char="@", header=FALSE)
names(x_tra) <- c("Displacement", "Horse_power", "Weight",
                 "Acceleration", "Model_year", "Mpg")
names(x_tst) <- c("Displacement", "Horse_power", "Weight",
                 "Acceleration", "Model_year", "Mpg")
if (tt == "train") { 
  test <- x_tra 
  }
else { 
  test <- x_tst 
  }
fitMulti=kknn(Mpg~Weight*Model_year + 
            Weight + I(Weight^2) + 
            Model_year + I(Model_year^2), train = x_tra, test = test)
yprime=fitMulti$fitted.values
sum(abs(test$Mpg-yprime)^2)/length(yprime) ##MSE
}
knnMSEtrain<-mean(sapply(1:5,run_knn_fold,name,"train"))
knnMSEtest<-mean(sapply(1:5,run_knn_fold,name,"test"))
cat("K-NN train score: ", knnMSEtrain, "\n")
cat("K-NN test score: ", knnMSEtest, "\n")
```
  Train and test scores for **Linear Model** and **k-NN**:

```{r}
data.frame(model_type = c("LM", "k-NN"), 
           train_score = c(lmMSEtrain, knnMSEtrain),
           test_score = c(lmMSEtest, knnMSEtest))
```
  
  k-NN achieves much lower training error than multiple linear regression, indicating better fit, but on test partitions its RMSE (8.76) is slightly higher than the linear model (8.56).
  Therefore, k-NN does not improve predictive performance on new data and shows a slight tendency to overfit compared to linear regression.

## 4.	Comparison of Multiple Regression Algorithms

  Read the CSV files and update LM and k-NN results for the autoMPG6 dataset:

```{r}
# read train file
res_train <- read.csv("../data/regr_train_alumnos.csv")

# update results of both models for autoMPG6
idx <- which(rownames(res_train) == "autoMPG6")  
res_train[res_train[,1] == "autoMPG6", "out_train_lm"]   <- lmMSEtrain
res_train[res_train[,1] == "autoMPG6", "out_train_kknn"] <- knnMSEtrain
write.csv(res_train, "../data/regr_train_alumnos.csv", row.names = FALSE)

tablatra <- as.matrix(res_train[, c("out_train_lm","out_train_kknn","out_train_m5p")])
colnames(tablatra) <- c("LM","KNN","M5")
rownames(tablatra) <- res_train[,1]

# same with the test file
res_test <- read.csv("../data/regr_test_alumnos.csv")

res_test[res_test[,1] == "autoMPG6", "out_test_lm"]   <- lmMSEtest
res_test[res_test[,1] == "autoMPG6", "out_test_kknn"] <- knnMSEtest
write.csv(res_test, "../data/regr_test_alumnos.csv", row.names = FALSE)

tablatst <- as.matrix(res_test[, c("out_test_lm","out_test_kknn","out_test_m5p")])
colnames(tablatst) <- c("LM","KNN","M5")
rownames(tablatst) <- res_test[,1]

```

  To compare LM and k-NN across all datasets, the Wilcoxon test was applied: 

```{r}
# Calculate relative difference LM - KNN
difs <- (tablatst[, "LM"] - tablatst[, "KNN"]) / tablatst[, "LM"]

# Transform data for the Wilcoxon test
# Create two columns containing negative and positive differences
wilc_1_2 <- cbind(ifelse(difs < 0, abs(difs) + 0.1, 0 + 0.1), # LM is worse than KNN
                  ifelse(difs > 0, abs(difs) + 0.1, 0 + 0.1)) # LM is better than KNN
colnames(wilc_1_2) <- c("LM","KNN")

# Two-sided paired Wilcoxon test LM vs KNN
LMvsKNNtst <- wilcox.test(wilc_1_2[,1], wilc_1_2[,2],
                          alternative = "two.sided",
                          paired = TRUE)
# Save R+ and p-value from the first run
Rmas   <- LMvsKNNtst$statistic
pvalue <- LMvsKNNtst$p.value

# Repeat the test to get the R- statistic
LMvsKNNtst <- wilcox.test(wilc_1_2[,2], wilc_1_2[,1],
                          alternative = "two.sided",
                          paired = TRUE)
Rmenos <- LMvsKNNtst$statistic

# Display result
data.frame(R_plus = Rmas, R_minus = Rmenos, p_value = pvalue)
```

  The p-value obtained is 0.966, so the null hypothesis of equal performance is not rejected. LM and k-NN show no statistically significant differences.
  
  Using the Friedman test to simultaneously compare all three algorithms (LM, k-NN, and M5'):

```{r}
# Run Friedman test
test_friedman <- friedman.test(as.matrix(tablatst))
test_friedman

# Prepare pairwise Wilcoxon test with Holm correction
tam <- dim(tablatst)
groups <- rep(1:tam[2], each = tam[1])

# Apply Holm post-hoc
pairwise.wilcox.test(as.matrix(tablatst),groups,p.adjust = "holm",paired  = TRUE)
```
  The Friedman test on LM, k-NN, and M5' errors gives a p-value of 0.006 < 0.05, indicating significant differences between at least one pair of algorithms. However, post-hoc analysis using paired Wilcoxon comparisons with Holm correction shows p-values of 0.832 for LM/k-NN and 0.062 for both LM/M5' and k-NN/M5' pairs, so no statistically significant differences between specific pairs are detected.
  
  All error results for autoMPG6 by the three algorithms:

```{r}
tablatst["autoMPG6", ]
```
  According to the results, the M5' algorithm achieves slightly lower error than the other two. Therefore, all three algorithms show comparable predictive performance.
